---
title: "Date and Time in R"
output: html_notebook
---
Being a Short Training Material used at Offa R Users Group Meeting on 29th July, 2024 at Statistical Laboratory, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria

Prepared and Presented
 By
Udokang, Anietie Edem (Oganizer, Orug)
Chief Lecturer, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria

###Introduction
Many data sets contains dates and day time which are not in form of a more familiar string or numerical data. This presentation shall be in two sections to address time in R and dates/time classes in R.
###Section 1
##Time in R
#What is the Current Time?
We can use R program to check current time by using Sys.time function.
```{r}
(now_ct <- Sys.time ())
```
#Time zones in R
In working with time and dates we will make use of R package lubridate.
```{r}
library(lubridate)
time_zones <- OlsonNames()
print(time_zones)
```
#Displays different time zones in R that should be used when converting from one time zone to another
#Change time zone
```{r}
time_Nigeria <- ymd_hms("2024-07-29 02:33:47", tz = "Africa/Lagos")
time_ny <- with_tz(time_Nigeria, tzone = "America/New_York")
time_ny
```
###Section 2
##DATES/TIME CLASSES IN R
The three classes are Date, POSITct and POSITlt’
##Date 
The function as.Date converts date-time objects to the Date class, which only stores the date (year, month, day) without the time. It’s useful when you only need the date part of a date-time object.
```{r}
date <- as.Date("2024-07-29")
print(date)
```
##Date with Day, Month, Year
```{r}
date <- as.Date("2021-01-25")
formatted_date <- format(date, format="%m/%d/%y")
print(formatted_date)  
```
##Format Date as Weekday:
```{r}
date <- as.Date("2021-01-25")
formatted_date <- format(date, format="%A")
print(formatted_date) 
```
##Format Date as Month:
```{r}
date <- as.Date("2021-01-25")
formatted_date <- format(date, format="%B")
print(formatted_date)
```
###POSITct
This function as.POSTct converts date-time objects to the POSIXct class, which stores date-time values as the number of seconds since January 1, 1970 (the Unix epoch). It’s useful for efficient storage and computation.
```{r}
datetime <- as.POSIXct("2024-07-29 03:17:16")
print(datetime)
```
##Date and Time:
```{r}
datetime <- as.POSIXct("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%Y-%m-%d %H:%M:%S")
print(formatted_datetime)  
```
##Custom Date and Time Format:
```{r}
datetime <- as.POSIXct("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%d-%b-%Y %I:%M %p")
print(formatted_datetime)  
```
##POSITlt
This function as.POSITlt converts date-time objects to the POSIXlt class, which stores date-time values as a list with components for seconds, minutes, hours, days, months, years, etc. It’s more detailed but less efficient for storage compared to POSIXct.
```{r}
datetime <- as.POSIXlt("2024-07-29 03:17:16")
print(datetime)
```
##Date and Time:
```{r}
datetime <- as.POSIXlt("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%Y-%m-%d %H:%M:%S")
print(formatted_datetime)  
```
##Custom Date and Time Format:
```{r}
datetime <- as.POSIXlt("2021-01-25 14:30:00")
formatted_datetime <- format(datetime, format="%d-%b-%Y %I:%M %p")
print(formatted_datetime)  
```
Thank you for spending your time with us.
Hope to see you in the next meeting.