---
title: "Design and Analysis OF Experiment with R"
output: html_notebook
---

Being a Short Training Material used at Offa-R –Users-Group Meeting On 27th June, 2024 at Statistical Laboratory, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria
Prepared By
Udokang, Anietie Edem (Oganizer, ORUG)
Chief Lecturer, Statistics Department, The Federal Polytechnic Offa, Kwara State, Nigeria
##Introduction
You are welcome to Offa-R-Users-Group (ORUG) for the June meeting here in the Statistical Laboratory, Statistics Department, The Federal Polytechnic Offa. Today we want to discuss how to effectively use R Statistical Software in design and analysis of experiment. In order to achieve this the discussion will be based on understanding the meaning, use and concept experimental design and thereafter, the application/use of R in Design and Analysis of Experiment.

##Design and Analysis of Experiment
Design an analysis of experiment is the process of planning an experiment, collecting data from such experiment and use the data collected for analysis for a valid inference about any problem under investigation. The experiment is designed in such a way that every experimental units has equal probability of receiving any treatment. This process useful in analyzing differences among group means caused by some factors.
Design an analysis of experiment has three basic principles which are replication, blocking and randomization.

##The Basic Principles of Design and Analysis of Experiment
 
#Replication
This is when the same experiment is repeated two or more times to verify results and brings about accuracy. If a treatment is administered to an experimental unit more than once that is referred to as replication.
#Randomization
This is the process of assigning the treatment to experimental unit randomly in such a way that every experimental units has equal chance or probability of receiving any treatment. This can practically be done by using lottery technique, table of random numbers, computer, etc.
#Blocking
This is the arrangement of experimental units into groups of the same characteristics or variables in order to reduce to minimum the impact of their variability on the observed outcomes. 

##Some Basic Terms in Experimental Design
#Experimental Units
These are the smallest units where treatment is being applied. For example a substance (treatment) is injected to each of the fish in an aquarium (plot). Each of the fish is the experimental unit. The plot in the experiment can as well be referred to as experimental unit if the treatment is applied on the whole plot not on each of the units in the plot. For example, the substance in the water (treatment) is applied to an aquarium. In this case the aquarium can be referred to as experimental unit.
#Sampling/Observational Units 
These are the units by which the responses are measured which may not be the same thing as experimental unit.
#Factors
These are variables of study that influence the response in an experiment. For example, feeds given to poultry birds. The factors can be discrete or continuous. For every factor there must be levels in an experiment. In the example given earlier, the different types of feeds forms the levels. Factors can also be classified as fixed and random, it is fixed if all the factors are included in the experiment and random if all the factors are not included in the experiment but are randomly selected.
#Treatments
These are different procedures/objects or factors that their effects on the experimental units are to be compared.

##Types of Experimental Design and its Model
The experimental design models are Completely Randomized Design (CRD), Completely Randomized Block Design (CRBD) and Factorial Design.
But due to limitation on time, the discussion today will be on CRD.

##Completely Randomized Design
#Definition
This is an experiment where all treatments are randomly allocated to the experimental units
Model:$Y_i_j = \mu + \alpha + \epsilon_i_j$
$i=1,2,3,...,n$  
$j=1,2,3,...,n$  

Where 
$\mu$ is the grand mean response
$\alpha$ is the effect due to the  level of the factor
$\epsilon$ is the random error

##Hypotheses
$H_0$: The effects of the levels of the factor on the response are significantly the same
Vs
$H_1$: At least one of the levels of the factor has a significant effect on the response

##Ilustrative Example
In an experiment, assuming we have three homogeneous chicken pens (plots) for three types of feeds. There are 21 broilers having same characteristics at the start of experiment with 7 broilers in each plot. The objective of the experiment is to determine the effects of three different composition of feeds on the growth rate of broilers. Let’s also assume that the experiment satisfied all other conditions to be referred to as a Completely Randomized Design (CRD). 
Table 1: The average growth rate (%) of the 21 broilers for seven weeks in three plots with three types of feeds.

Wk F1 F2  F3
1	20	21	22
2	30	20	23
3	23	23	24
4	21	25	21
5	22	21	23
6	25	30	22
7	26	19	20

```{r}
response = c(20,21,22,30,20,23,23,23,24,21,25,21,22,21,23,25,30,22,26,19,20)
response
```
```{r}
factor = c("FI", "FII", "FIII")
k=3 #number of levels
n=7 #number of items under each treatment
tm = gl(k, 1, n*k, factor(factor)) #treatment matching of levels that corresponds to each response
tm
```
```{r}
crd = aov(response ~ tm)
print(crd)
```
```{r}
summary(crd)
```

#If the response data is the excel in matrix form
```{r}
Broiler = read.csv("C:/Users/FEDPOFFA/Documents/Broiler.csv", header=TRUE)
Broiler
```
```{r}
response = c(t(as.matrix(Broiler)))#response data in a single vector
response
```


```{r}
factor = c("FeedI", "FeedII", "FeedIII")
k=3
n=7
tm = gl(k, 1, n*k, factor(factor)) #treatment matching of levels that corresponds to each response
tm
```
```{r}
crd = aov(response ~ tm)
print(crd)
```
```{r}
summary(crd)
```
#Comment: 
The p-value = 0.573 lead to the acceptance of the null hypothesis, hence the three types feeds does not have any significant different effect on the growth rate of broilers.

##6. R Packages for Analysis of Design of Experiment
Some of these packages are agricolae,  doebioresearch, ibd and augmentedRCBD.

THANK YOU FOR PARTICIPATION